---
title: "r-week-8-hw-eeb174"
author: "Mehdi Nojoumi"
date: "March 5, 2017"
output: md_document
---

# Family Canids:
***
<br><br>
```{r , echo=TRUE}
#install.packages("paleotree")
library(ggplot2)
canids <- read.csv("~/Desktop/eeb-177/homework/canids-ranges.csv", header = F, as.is = T)
names(canids) <- c("genus", "species", "minage", "maxage")
head(canids)

canid_occ <- ggplot(canids, aes( species, ymin = maxage, ymax=minage, colour = genus))
canid_occ <- canid_occ + geom_linerange()+ theme(legend.position="none")+ coord_flip()+  theme(axis.text.y = element_text(size=3))+ theme(axis.ticks.y=element_blank()) + scale_y_continuous(limits=c(0, 40), expand = c(0, 0), breaks=c(0, 10, 20, 30, 40)) + labs(title = "Canid Fossil Occurrences", x = "Species", y = "Ma ago") + theme(plot.title = element_text(hjust = 0.5, size=22, face = "bold"), axis.title =element_text(size=20))


canid_occ

ggsave(filename = "canid-occ.pdf", plot = canid_occ)

```

# Report for family Canids:
***
1.What taxonomic group is displayed?
 <br> Figures are for the group Canids. 

2.How many unique fossil species are contained within it? 
<br> I've used the following command <br>
cat canids-ranges.csv | cut -d "," -f 2 | sort | uniq | wc -l <br>
Answer: There are 160 unique fossil species.


3.How many unique fossil genera? 
<br> I've used the following command
<br> cat canids-ranges.csv | cut -d "," -f 1 | sort | uniq | wc -l
<br> Answer: There are 58 uniqe genera.

4.What species has the longest(range)?
<br> Longest fossil range belongs to Cynarctoides acridens , with the fossil range =  [17.200000000000003]



<br><br><br>

# Family Cetacea:
***



```{r , echo=TRUE}
#install.packages("paleotree")
library(ggplot2)
cetacea <- read.csv("~/Desktop/eeb-177/homework/cetacea-ranges.csv", header = F, as.is = T)
names(cetacea) <- c("genus", "species", "minage", "maxage")
head(cetacea)

cetacea_occ <- ggplot(cetacea, aes( species, ymin = maxage, ymax=minage, colour = genus))
cetacea_occ <- cetacea_occ + geom_linerange()+ theme(legend.position="none")+ coord_flip()+  theme(axis.text.y = element_text(size=3))+ theme(axis.ticks.y=element_blank()) + scale_y_continuous(limits=c(0, 40), expand = c(0, 0), breaks=c(0, 10, 20, 30, 40)) + labs(title = "Cetacea Fossil Occurrences", x = "Species", y = "Ma ago") + theme(plot.title = element_text(hjust = 0.5, size=22, face = "bold"), axis.title =element_text(size=20))


cetacea_occ

ggsave(filename = "cetacea-occ.pdf", plot = cetacea_occ)


```

# Report for family Canids:
***
1.What taxonomic group is displayed?
 <br> Figures are for the group Cetacea. 

2.How many unique fossil species are contained within it? 
<br> I've used the following command <br>
cat cetacea-ranges.csv | cut -d "," -f 2 | sort | uniq | wc -l <br>
Answer: There are 545 unique fossil species.


3.How many unique fossil genera? 
<br> I've used the following command
<br> cat cetacea-ranges.csv | cut -d "," -f 1 | sort | uniq | wc -l
<br> Answer: There are 354 uniqe genera.

4.What species has the longest(range)?
<br> Longest fossil range belongs to two species : 
Longest fossil range belongs to Eubalaena australis , with the fossil range =  [23.03]
Longest fossil range belongs to Megaptera novaeangliae , with the fossil range =  [23.03]









